Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-20 writeup/conStruct.tex
added script for plosgen submission

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

%% added PLR
\newif\ifsubmissionversion
%\submissionversionfalse
\submissionversiontrue
\newif\ifincludesupplement
\includesupplementtrue
%\includesupplementfalse
\newif\iffloatsatend
\floatsatendfalse
%\floatsatendtrue

\iffloatsatend
\usepackage{endfloat}
\fi
%% end added PLR

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref}
\usepackage[hidelinks]{hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\ifsubmissionversion
    \raggedright
    \setlength{\parindent}{0.5cm}
    \textwidth 5.25in 
    \textheight 8.25in
\else
    \usepackage{fullpage}
\fi

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{latex/plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{graphicx}
\ifsubmissionversion
    \usepackage{lastpage,fancyhdr}
    \usepackage{epstopdf}
    \pagestyle{myheadings}
    \pagestyle{fancy}
    \fancyhf{}
    \setlength{\headheight}{27.023pt}
    \lhead{\includegraphics[width=\textwidth]{latex/PLOS-submission.eps}}
    \rfoot{\thepage/\pageref{LastPage}}
    \renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
    \fancyheadoffset[L]{2.25in}
    \fancyfootoffset[L]{2.25in}
    \lfoot{\sf PLOS}
\fi

@@

\newcommand{\e}[1]{{\mathbb E}\left[ #1 \right]}
\newcommand{\given}{\mid}
\newcommand{\secref}[1]{``\nameref{#1}''}
\newcommand{\tri}{\textit{trichocarpa}} 
\newcommand{\bals}{\textit{balsamifera}}

\newcommand{\gb}[1]{{\it\color{magenta}{(#1)}}}
\newcommand{\plr}[1]{{\it\color{purple}{(#1)}}}
\newcommand{\gc}[1]{{\it\color{blue}{(#1)}}}

\newcommand{\R}{\mathbb{R}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\supp}{\mathop{\mbox{supp}}}
\newcommand{\sgn}{\mathop{\mbox{sgn}}}
\newcommand{\bone}{\mathbf{1}}

@@

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Inferring continuous and discrete population genetic structure across space}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Gideon S.\  Bradburd\textsuperscript{1,a},
Graham M.\  Coop\textsuperscript{2,b,\Yinyang},
Peter L.\  Ralph\textsuperscript{3,c,\Yinyang},
\\
\bigskip
\textbf{1}
Department of Integrative Biology, 
Ecology, Evolutionary Biology, and Behavior Graduate Group,
Michigan State University, MI 48824
\\
\textbf{2}
Center for Population Biology,
Department of Evolution and Ecology, 
University of California, Davis, CA 95616
\\
\textbf{3} 
Institute of Ecology and Evolution,
Departments of Mathematics and Biology,
University of Oregon, Eugene, OR 97403
\\
\textbf{a} bradburd@msu.edu; 
\textbf{b} gmcoop@ucdavis.edu;
\textbf{c} plr@uoregon.edu

@@

%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This thereby explicitly addresses the ``clines versus clusters" problem in 
modeling population genetic variation.
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.

\ifsubmissionversion
\newpage
\fi

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
One of the first steps in the analysis of genetic data, 
and a principal mission of biology, 
is to describe and categorize natural variation. 
A continuous pattern of differentiation (isolation by distance), 
where individuals found closer together in space 
are, on average, more genetically similar than individuals sampled farther apart, 
can confound attempts to categorize natural variation into groups.  
This is because current statistical methods for assigning individuals to discrete clusters 
cannot accommodate spatial patterns, 
and so are forced to use clusters to describe what is in fact
continuous variation. 
As isolation by distance is common in nature, 
this is a substantial shortcoming of existing methods.
In this study, we introduce a new statistical method 
for categorizing natural genetic variation - 
one that describes variation as a combination of continuous and discrete patterns.
We demonstrate that this method works well and can capture patterns in
population genomic data without resorting to splitting populations
where they can be described by continuous patterns of variation.

%Using simulated data, we show how current approaches can be misleading, 
%and also demonstrate that our new method works well.  
%We predict that our method will be an important tool for the analysis of genetic data.

\ifsubmissionversion
\linenumbers
\fi

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \cite{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

\emph{Isolation by distance} (IBD) refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\cite{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\cite{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\cite{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \cite{barton2002neutral,barton2013modelling}.
% Isolation by distance has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \cite{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that isolation by distance is very widespread in nature \cite{meirmans2012,Sexton_etal_2014}.

The ubiquity of isolation by distance presents a challenge for models of discrete population structure,
as it is frequently difficult to determine whether observed patterns of genetic variation are 
continuously distributed across a landscape, or instead are partitioned in discrete clusters.
This problem can be compounded if sampling is done unevenly or discretely across a population or species' range,
and has given rise to a debate in the population genetic literature
about how best to describe sets of individuals using continuous clines and discrete clusters 
(e.g., \cite{SerrePaabo2004,rosenberg2005clines}).

Existing model-based clustering approaches can only describe continuous patterns of variation using
discrete clusters, and so tend to erroneously describe continuously distributed variation with multiple clusters that 
show spatially autocorrelated cluster membership \cite{Frantz2009,meirmans2012}.
In analyses of empirical datasets, which often show strong isolation by distance,
model-based clustering approaches will therefore tend to overestimate
the number of discrete clusters present.

To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.
In this paper, 
% we present a statistical method for simultaneously 
% modeling continuous and discrete patterns of population structure.
we model genetic variation in genotyped individuals as 
partitioned within or admixed across a specified number of discrete layers,
within each of which relatedness decays as a parametric function of the distance between samples.
We also implement a cross-validation approach for comparing and selecting models across different numbers of layers,
and we demonstrate the utility of our approach using both simulated and empirical data.
The implementation of this method, \texttt{conStruct} (for ``\emph{con}tinuous \emph{struct}ure"), 
is available as an R package at 
\href{https://github.com/gbradburd/conStruct}{https://github.com/gbradburd/conStruct}.

% \section*{Methods}
\section*{Results}

>>>>

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %
\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
%% added PLR
\newif\ifsubmissionversion
%\submissionversionfalse
\submissionversiontrue
\newif\ifincludesupplement
\includesupplementtrue
%\includesupplementfalse
\newif\iffloatsatend
\floatsatendfalse
%\floatsatendtrue
\newif\ifincludefigs
%\includefigstrue
\includefigsfalse
\iffloatsatend
\usepackage[nomarkers]{endfloat}
\fi
%% end added PLR
% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}
% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}
% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}
% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref}
\usepackage[hidelinks]{hyperref}
% line numbers
\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}
% array package and thick rules for tables
\usepackage{array}
% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}
% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}
% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}
% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing
% Text layout
\ifsubmissionversion
    \raggedright
    \setlength{\parindent}{0.5cm}
    \textwidth 5.25in 
    \textheight 8.25in
\else
    \usepackage{fullpage}
\fi
% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}
% Use the PLoS provided BiBTeX style
\bibliographystyle{latex/plos2015}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother
% Leave date blank
\date{}
% Header and Footer with logo
\usepackage{graphicx}
\ifsubmissionversion
    \usepackage{lastpage,fancyhdr}
    \usepackage{epstopdf}
    \pagestyle{myheadings}
    \pagestyle{fancy}
    \fancyhf{}
    \setlength{\headheight}{27.023pt}
    \lhead{\includegraphics[width=\textwidth]{latex/PLOS-submission.eps}}
    \rfoot{\thepage/\pageref{LastPage}}
    \renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
    \fancyheadoffset[L]{2.25in}
    \fancyfootoffset[L]{2.25in}
    \lfoot{\sf PLOS}
\fi

@@

\newcommand{\e}[1]{{\mathbb E}\left[ #1 \right]}
\newcommand{\given}{\mid}
\newcommand{\secref}[1]{``\nameref{#1}''}
\newcommand{\tri}{\textit{trichocarpa}} 
\newcommand{\bals}{\textit{balsamifera}}
\newcommand{\gb}[1]{{\it\color{magenta}{(#1)}}}
\newcommand{\plr}[1]{{\it\color{purple}{(#1)}}}
\newcommand{\gc}[1]{{\it\color{blue}{(#1)}}}
\newcommand{\R}{\mathbb{R}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\supp}{\mathop{\mbox{supp}}}
\newcommand{\sgn}{\mathop{\mbox{sgn}}}
\newcommand{\bone}{\mathbf{1}}

@@

\begin{document}
\vspace*{0.2in}
% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Inferring continuous and discrete population genetic structure across space}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Gideon S.\  Bradburd\textsuperscript{1,a},
Graham M.\  Coop\textsuperscript{2,b,\Yinyang},
Peter L.\  Ralph\textsuperscript{3,c,\Yinyang},
\\

@@

\textbf{1}
Department of Integrative Biology, 
Ecology, Evolutionary Biology, and Behavior Graduate Group,
Michigan State University, MI 48824
\\
\textbf{2}
Center for Population Biology,
Department of Evolution and Ecology, 
University of California, Davis, CA 95616
\\
\textbf{3} 
Institute of Ecology and Evolution,
Departments of Mathematics and Biology,
University of Oregon, Eugene, OR 97403
\\
\textbf{a} bradburd@msu.edu; 
\textbf{b} gmcoop@ucdavis.edu;
\textbf{c} plr@uoregon.edu
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This thereby explicitly addresses the ``clines versus clusters" problem in 
modeling population genetic variation.
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.
\ifsubmissionversion
\newpage
\fi
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
One of the first steps in the analysis of genetic data, 
and a principal mission of biology, 
is to describe and categorize natural variation. 
A continuous pattern of differentiation (isolation by distance), 
where individuals found closer together in space 
are, on average, more genetically similar than individuals sampled farther apart, 
can confound attempts to categorize natural variation into groups.  
This is because current statistical methods for assigning individuals to discrete clusters 
cannot accommodate spatial patterns, 
and so are forced to use clusters to describe what is in fact
continuous variation. 
As isolation by distance is common in nature, 
this is a substantial shortcoming of existing methods.
In this study, we introduce a new statistical method 
for categorizing natural genetic variation - 
one that describes variation as a combination of continuous and discrete patterns.
We demonstrate that this method works well and can capture patterns in
population genomic data without resorting to splitting populations
where they can be described by continuous patterns of variation.
%Using simulated data, we show how current approaches can be misleading, 
%and also demonstrate that our new method works well.  
%We predict that our method will be an important tool for the analysis of genetic data.
\ifsubmissionversion
\linenumbers
\fi
\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.
From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \cite{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.
There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.  
\emph{Isolation by distance} (IBD) refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\cite{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\cite{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\cite{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \cite{barton2002neutral,barton2013modelling}.
% Isolation by distance has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \cite{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that isolation by distance is very widespread in nature \cite{meirmans2012,Sexton_etal_2014}.
The ubiquity of isolation by distance presents a challenge for models of discrete population structure,
as it is frequently difficult to determine whether observed patterns of genetic variation are 
continuously distributed across a landscape, or instead are partitioned in discrete clusters.
This problem can be compounded if sampling is done unevenly or discretely across a population or species' range,
and has given rise to a debate in the population genetic literature
about how best to describe sets of individuals using continuous clines and discrete clusters 
(e.g., \cite{SerrePaabo2004,rosenberg2005clines}).
Existing model-based clustering approaches can only describe continuous patterns of variation using
discrete clusters, and so tend to erroneously describe continuously distributed variation with multiple clusters that 
show spatially autocorrelated cluster membership \cite{Frantz2009,meirmans2012}.
In analyses of empirical datasets, which often show strong isolation by distance,
model-based clustering approaches will therefore tend to overestimate
the number of discrete clusters present. 
To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.
In this paper, 
% we present a statistical method for simultaneously 
% modeling continuous and discrete patterns of population structure.
we model genetic variation in genotyped individuals as 
partitioned within or admixed across a specified number of discrete layers,
within each of which relatedness decays as a parametric function of the distance between samples.
We also implement a cross-validation approach for comparing and selecting models across different numbers of layers,
and we demonstrate the utility of our approach using both simulated and empirical data.
The implementation of this method, \texttt{conStruct} (for ``\emph{con}tinuous \emph{struct}ure"), 
is available as an R package at 
\href{https://github.com/gbradburd/conStruct}{https://github.com/gbradburd/conStruct}.
% \section*{Methods}
\section*{Results}
\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
We write $N_i$ for the number of chromosomes sequenced in the $i^\text{th}$ sample.
The \emph{sample frequency} at locus $\ell$ in sample $n$, denoted $f_{n,\ell}$, 
is calculated by first arbitrarily choosing one of the observed alleles at locus $\ell$ to count, 
then dividing the number of observations of that counted allele by the total number of chromosomes genotyped at that locus
in sample $n$.
The choice of allele does not affect subsequent calculations, and so may be arbitrary.
We then calculate the \emph{allelic covariance} between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles 
chosen from each of the two samples at a random locus,
coding the counted allele as `1' (and the other as `0').
Concretely, this is
\begin{equation}
\widehat{\Omega}_{i,j} = 
    \frac{1}{L} \sum_{\ell=1}^L \left( (f_{i,\ell}-1/2) (f_{j,\ell}-1/2) - \delta_{i,j} f_{i,\ell}^2 / (N_i-1) \right),
\label{allelic_covariance}
\end{equation}
where $\delta_{i,j}=1$ if $i=j$ and is 0 otherwise.
Although we describe this as a covariance between individually drawn alleles,
$\widehat{\Omega}_{i,j}$ is in fact also the covariance between the allele frequencies
of a randomly chosen allele in samples $i$ and $j$, as long as $i \neq j$.
The diagonal (where $i=j$) does not have this interpretation, 
and reflects the covariance between individual alleles drawn from within the population.
This definition of covariance differs from the usual ``genetic covariance'' \cite{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choice of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eq.\ \eqref{allelic_covariance} further in \secref{allelic_cov}.
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
